---
title: "DevMethyl"
author: "Hannah Glaser"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DevMethyl}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r message=FALSE, warning=FALSE}
library(DevMethyl)
```

```{r}
load_exdata() -> file_paths
file_paths[[1]] -> methylpath
file_paths[[2]]-> accpath
system.file("extdata", "column_header.txt", package = "DevMethyl") -> headerpath
system.file("extdata", "column_header_accessibility.txt", package = "DevMethyl") -> acc_headerpath
system.file("extdata", "cell-annotation.tsv", package = "DevMethyl") -> metapath
system.file("extdata", "cpgIslandExtUnmasked.txt" , package = "DevMethyl") -> cpgipath
```

## 2.2 Preparations
The analysis starts by choosing a genomic region of interest (ROI). Depending on the extend of your analysis it might save you time to assign your settings to variables. In this example we chose chromosome 8 to analyse the variable region (VR) 8653165-8659055. To compare our VR with non-VRs we chose our ROI 8628165-8684055. If you don't know any VRs you can leave these variables out and continue using a random ROI. Depending on the used sparse matrix and ROI, you might need to adjust your kernel size (hx, ht).
```{r Settings}
chr <- 8
rstart <- 8628165
rend <- 8684055
startVR<- 8653165
endVR <- 8659055
hx <- 200
ht <- 0.4
```

# 3 Visualizing methylation data
If you are working with a numpy array file (npz), you have to convert it into a sparse matrix first using `npz.to.spM`.
```{r npz.to.spM()}
npz.to.spM(methylpath) -> sp
```

To plot your methylation data, you first use `map_methyl()` to map your sparse matrix to the cell ids and pseudotimes.
For this you have to make sure the cell ids are stored in a column "cell_id" of a data frame and are of the same format as in your meta data containing the pseudotimes.
```{r}
header <- readLines(headerpath) %>% as.data.frame()

names(header)[names(header) == "."] <- "cell_id"
  
header %>%
  mutate(cell_id = sub("_(?!A).*", "",cell_id, perl = TRUE)) -> header

```

Your meta data need to store the cell ids in the column "cell_id_dna" and the pseudotimes in "pt_avg".
Depending on the amount of pseudotimes available for your cells you might need to take average values. 
```{r}
meta <- read.table(metapath, header =TRUE, sep = "\t")
   meta$pt_avg <- rowMeans(meta[44:45], na.rm = TRUE)
```

`map_methyl()` will return a data frame containing the methylation status (methyl) of the genomic position (pos) for each pseudotime (pt).
```{r map_methyl()}
map_methyl(sp, meta, header, rstart, rend) -> mappedpt
mappedpt %>% head()
```

With the mapped pseudotimes and your chosen kernel size, you now smooth the data using `smooth2d`. The function will return a matrix with your values smoothed by the a gaussian kernel.
```{r gauss_kernel}
smooth2d(mappedpt, hx, ht) -> m
```

Finally you can plot your matrix using `plot_methyl`. This will return a tile plot visualizing how the methylation pattern changes during differentiation.
```{r plot_methyl, fig.width=7}
plot_methyl(m, mappedpt, rstart)
```

The same can be repeated for chromatin accessibility data.
```{r accessibility, echo=FALSE}
npz.to.spM(accpath) ->spAcc

header_acc <- readLines(acc_headerpath) %>% as.data.frame()
 
names(header_acc)[names(header_acc) == "."] <- "cell_id"

  header_acc %>%
    mutate(cell_id = sub("_(?!A).*", "",cell_id, perl = TRUE)) -> header_acc
  map_methyl(spAcc, meta, header_acc, rstart, rend) -> mappedpt_acc
  smooth2d(mappedpt_acc, hx, ht) -> mAcc
```

```{r plot_methyl(access), fig.width=7}
plot_methyl(mAcc, mappedpt_acc, rstart)
```


# 4 Visualizing genomic attributes
To get a comprehensive outlook of your ROI, you can look at several genomic attributes using this package. 

## 4.1 CpG islands
With `plot_cpgislands` you can plot your data table retrieved from https://genome.ucsc.edu/cgi-bin/hgTables to visualize CpG islands within your ROI.
```{r plot_CpGi, fig.width=7}
plot_cpgislands(cpgipath, chr, rstart, rend)
```

## 4.2 CpG density
Further you can look at the CpG distribution in you ROI. `get_cpgs` returns a integer list containing the genomic positions of the CpGs.
```{r}
get_cpgs(chr, rstart, rend) %>% head()
```

This can be plotted as a bar plot using `plot_cpgs`.
```{r plot_CpGs, fig.width=7}
plot_cpgs(chr, rstart, rend)
```

## 4.3 Genemodel
It can be especially relevant to compare methylation patterns to genes and genomic features. 
For this you need to extract the genome dataset of your chosen species from the ensemble FTP site. The function `get_genemodel` returns a data frame containing genomic information like start and end point of genes, exon, coding regions and more.
```{r genemodel, fig.width=12}
genepath<- "https://ftp.ensembl.org/pub/release-110/gtf/mus_musculus/Mus_musculus.GRCm39.110.gtf.gz"
get_genemodel(genepath, chr, rstart, rend) %>% head(3)
```

To plot the genemodel use `plot_genemodel`, which returns a arrow plot of genes including colored annotations of the genomic features. You can either store the file and use the directory or you can access the file using the url.
```{r plot genemodel, fig.width=7}
plot_genemodel(genepath, chr, rstart, rend)
```

## 4.4 Regulartory feature
To get the regulatory features, extract the regulation dataset of your chosen species from the ensemble FTP site. The function `get_regulation` returns a data frame containing start and end of enhancers and promoters and more
```{r reg features}
regpath<- "https://ftp.ensembl.org/pub/release-110/regulation/mus_musculus/mus_musculus.GRCm39.Regulatory_Build.regulatory_features.20221007.gff.gz"
get_regulation(regpath, chr, rstart, rend) %>% head(1)
```

To plot the regulatory features use `plot_regulation`, which returns a segment plot including a colored annotation of the variable region if given . You can either store the file and use the directory or you can access the file using the url.
```{r fig.width=7}
plot_regulation(regpath, chr, rstart, rend, startVR, endVR)
```
  
# 5 Creating summary plot
The easiest way to get a comprehensive summary is to plot all data in one plot using `plot_all`. 
```{r plot features, fig.height=8, fig.width=10}
plot_all(cpgipath, methylpath, meta, header, accpath, header_acc, regpath, genepath, hx, ht, chr, rstart, rend, startVR, endVR)

```

# 6 Session Info
```{r Session Info}
sessionInfo()
```

